Population scale mapping of transposable element diversity reveals links to gene regulation and epigenomic variation

### 2016-08-12 Paper/paper.tex
Minor formatting changes
Changed font
Update text

@@

@@

Variation in the presence or absence of transposable elements (TEs) is a
major source of genetic variation between individuals. Here, we
identified 23,095 TE presence/absence variants between 216 wild
Arabidopsis accessions. Most variants are rare, and we find a burden of
rare variants associated with extremes of both gene expression and DNA
methylation levels within the population. Of the common alleles
identified, two thirds were not in linkage disequilibrium with nearby
SNPs, implicating these TE variants as a source of novel genetic
diversity. Nearly 200 common TE variants were associated with
significantly altered expression of nearby genes, with potential
functional consequences including decreased pathogen resistance. A major
fraction of inter-accession DNA methylation differences were associated
with nearby TE insertion variants, with DNA, indicating an important
role of TE variants in facilitating epigenomic variation.

@@

Transposable elements (TEs) are mobile genetic elements present in
nearly all studied organisms, and make up a large fraction of most
eukaryotic genomes. The two classes of TEs are retrotransposons (class
I elements), which transpose via an RNA intermediate requiring a
reverse transcription reaction, and DNA transposons (class II
elements), which transpose via either a cut-paste or, in the case of
Helitrons, a rolling circle mechanism with no RNA intermediate
\cite{Wicker:2007en}. TE activity poses mutagenic potential as a TE
insertion may disrupt essential regions of the genome. Consequently,
safeguards have evolved in order to suppress this activity. These
safeguards include epigenetic transcriptional silencing mechanisms,
chiefly involving the methylation of cytosine nucleotides (DNA
methylation) to produce 5-methylcytosine (mC), a mark that can signal
transcriptional silencing of the methylated locus. In
\emph{Arabidopsis thaliana }(Arabidopsis), DNA methylation occurs in
three DNA sequence contexts: mCG, mCHG, and mCHH, where H is any base
but G. Establishment of DNA methylation marks can be carried out by
two distinct pathways -- the RNA-directed DNA methylation pathway
guided by 24 nucleotide (nt) small RNAs (smRNAs), and the DDM1/CMT2
pathway \cite{Zemach:2013dj, Matzke:2014ek}. A major function of DNA
methylation in Arabidopsis is in the transcriptional silencing of TEs.
Loss of DNA methylation due to mutations in genes essential for DNA
methylation establishment and maintenance leads to expression of
previously silent TEs, and in some cases transposition
\cite{Mirouze:2009km, Miura:2001eg, Saze:2003du, Lippman:2004cm,
  Jeddeloh:1999gu, Zemach:2013dj}.

TEs are thought to play an important role in evolution, not only because
of the disruptive potential of their transposition. The release of
transcriptional and post-transcriptional silencing of TEs can lead to
bursts of TE activity, quickly generating new genetic diversity
\cite{Vitte:2014he}. TEs may carry regulatory information such as
promoters and transcription factor binding sites, and their mobilization
may lead to the creation or expansion of gene regulatory networks
\cite{Henaff:2014hg, Bolger:2014bn, Ito:2011dga, Makarevitch:2015ho}.
Furthermore, the transposase enzymes required and encoded by TEs have
frequently been domesticated and repurposed as endogenous proteins, such
as the \emph{DAYSLEEPER} gene in Arabidopsis, derived from a hAT
transposase enzyme \cite{Bundock:2005gp}. Clearly, the activity of TEs
can have widespread and unpredictable effects on the host genome.
However, the identification of TE presence/absence variants in genomes
has remained difficult to date. It is challenging to identify the
structural changes in the genome caused by TE mobilization using current
short-read sequencing technologies as these reads are typically mapped
to a reference genome, which has the effect of masking structural
changes that may be present. However, in terms of the number of base
pairs affected, a large fraction of genetic differences between
Arabidopsis accessions appear to be due to variation in TE content
\cite{Cao:2011cf, Quadrana:2016bi}. Therefore identification of TE
variants is essential in order to develop a more comprehensive
understanding of the genetic variation that exists between genomes, and
of the consequences of TE movement on genome and cellular function.

The tools developed previously for identification of novel TE insertion
sites have several limitations. They either require a library of active
TE sequences, cannot identify TE absence variants, or are not designed
with population studies in mind, or suffer from a high degree of
false-negatives
\cite{Thung:2014ir, Robb:2013gw, Henaff:2015dl, Quadrana:2016bi}. In
order to accurately map the locations of TE presence/absence variants
with respect to a reference genome, we have developed a novel algorithm,
TEPID (Transposable Element Polymorphism IDentification), which is
designed for population studies. We tested our algorithm using both
simulated and real Arabidopsis sequencing data, finding that TEPID is
able to accurately identify TE presence/absence variants with respect to
the Col-0 reference genome. We applied our TE analysis method to
existing genome resequencing data for 216 different wild Arabidopsis
accessions, and identified widespread TE variation amongst these
accessions \cite{Schmitz:2013iu}.

@@

In order to further assess the sensitivity and specificity of TE variant
discovery using TEPID, we identified TE variants in the \emph{Landsberg}
\emph{erecta} (L\emph{er}) accession, and compared these with the
L\emph{er }genome assembly created using long PacBio sequencing reads
\cite{Chin:2013iw}. Previously published 100 bp paired-end L\emph{er}
genome resequencing reads \cite{Schneeberger:2011ft} were first analyzed
using TEPID, enabling identification of 446 TE insertions (Figure 1 -
source data 1) and 758 TE absence variants (Figure 1 - source data 2)
with respect to the Col-0 reference TE annotation. Reads providing
evidence for these variants were then mapped to the L\emph{er} reference
genome that was generated by \emph{de novo} assembly using Pacific
Biosciences P5-C3 chemistry with a 20 kb insert library
\cite{Chin:2013iw}, using the same alignment parameters as was used to
map reads to the Col-0 reference genome. This resulted in 98.7\% of
reads being aligned concordantly to the L\emph{er} reference, whereas
100\% aligned discordantly or as split reads to the Col-0 reference
genome (Table \ref{table1}). To find whether reads mapped to homologous regions in
both the Col-0 and L\emph{er} reference genomes, we conducted a blast
search \cite{Camacho:2009fc} using the DNA sequence between read pair
mapping locations in the L\emph{er }genome against the Col-0 genome, and
found the top blast result for 80\% of reads providing evidence for TE
insertions, and 89\% of reads providing evidence for TE absence variants
in L\emph{er}, to be located within 200 bp of the TE variant
reported by TEPID. We conclude that reads providing evidence for TE
variants map discordantly or as split reads when mapped to the Col-0
reference genome, but map concordantly to homologous regions of the
L\emph{er} \emph{de novo} assembled reference genome, indicating that
structural variation is present at the sites identified by TEPID, and
that this is resolved in the \emph{de novo} assembled genome.

To estimate the rate of false negative TE absence calls made using
TEPID, we compared our L\emph{er} TE absence calls to the set of TE
absences in L\emph{er} genome identified previously by aligning
full-length Col-0 TEs to the Ler reference using BLAT
\cite{Quadrana:2016bi}. We found that 89.6\% (173/193) of these TE
absences were also identified using TEPID, indicating a false negative
rate of around 10\% for TE absence calls. To determine the rate of false
negative TE insertion calls, we ran TEPID using 90 bp paired-end Col-0
reads (Col-0 control samples from \cite{Jiang:2014ih}), aligning reads
to the L\emph{er} PacBio assembly. As TEPID requires a high-quality TE
annotation to discover TE variants, which is not available for the
L\emph{er} assembly, we simply looked for discordant and split read
evidence at the known Col-0-specific TE insertion sites
\cite{Quadrana:2016bi}, and found evidence reaching the TEPID threshold
for a TE insertion call to be made at 89.6\% (173/193) of these sites,
indicating a false negative rate of 10\%. However, we note that this
estimate does not take into account the TEPID refinement step used on
large populations, and so the false negative rate for samples analyzed
in the population from Schmitz et al. (2013) is likely to be slightly
lower than this estimate, as each accession gained on average 4\% more
insertion calls following this refinement step (Figure 2 - figure
supplement \hyperref[fig2s1]{1}).

@@

We used TEPID to analyze previously published 100 bp paired-end genome
resequencing data for 216 different Arabidopsis accessions
\cite{Schmitz:2013iu}, and identified 15,007 TE insertions and 8,088 TE
absence variants relative to the Col-0 reference accession, totalling
23,095 unique TE variants. In most accessions we identified 300-500 TE
insertions (mean = 378) and 1,000-1,500 TE absence variants (mean =
1,279), the majority of which were shared by two or more accessions
(Figure 2 - figure supplement \hyperref[fig2s2]{2}). PCR validations were performed for a
random subset of 10 insertions and 10 absence variants in 14 accessions
(totalling 280 validations), and confirmed the high accuracy of TE
variant discovery using the TEPID package, with results similar to that
observed using simulated data and the L\emph{er} genome analysis (Figure 2 -
figure supplement \hyperref[fig2s3]{3}). The number of TE insertions identified was
positively correlated with sequencing depth of coverage, while the
number of TE absence variants identified had no correlation with
sequencing coverage (Figure 2 - figure supplement \hyperref[fig2s4]{4}A, B), indicating
that the sensitivity of TE absence calls is not limited by sequencing
depth, while TE insertion calls benefitted from high sequencing depth.
However, accessions with low coverage gained more TE insertion calls
during the TEPID refinement step (Figure 2 - figure supplement \hyperref[fig2s4]{4}C),
indicating that these false negatives were effectively reduced by
leveraging TE variant information for the whole population.

As TE insertion and TE absence calls represent an arbitrary comparison
to the Col-0 reference genome, we sought to remove these arbitrary
comparisons and classify each variant as a new TE insertion or true
deletion of an ancestral TE in the population. To do this, we examined
the minor allele frequency (MAF) of each variant in the population,
under the expectation that the minor allele is the derived allele. We
re-classified common TE absences relative to Col-0 as rare TE insertions
in Col-0, and common TE insertions relative to Col-0 as true TE
deletions in Col-0. Cases where the TE variant was at an high MAF
(\textgreater{}20\%) were assigned NA calls, as it could not be
determined if these were cases where the variant was most likely to be
due to a true TE deletion or due to a new TE insertion. While these
classifications are not definitive, as there will be rare cases where a
true TE deletion has spread through the population and becomes the
common allele, we will correctly classify most TE variants. Overall, we
found 72.3\% of the TE absence variants identified with respect to the
Col-0 reference genome were likely due to a true TE deletion in these
accessions, while 4.8\% were due to new insertions in Col-0 not shared
by others in the population (Table \ref{table2}). Overall, we identified 15,077 new
TE insertions, 5,856 true TE deletions, and 2,162 TE variants at a high
MAF that were unable to be classified as an insertion or deletion.

TE insertions and deletions were distributed throughout chromosome 1 in
a pattern that was similar to the distribution of all Col-0 TEs (Figure
\ref{fig2}A). TE deletions and common TE variants were found in similar
chromosomal regions, as deletion variants represent the rare loss of
common variants. TE deletions and common variants were more highly
enriched in the pericentromeric regions than rare variants or TE
insertions. The distribution of rare TE variants and TE insertions was
somewhat similar to that observed for regions of the genome previously
identified as being differentially methylated in all DNA methylation
contexts (mCG, mCHG, mCHH) between the wild accessions (population
C-DMRs), while population CG-DMRs, differentially methylated in the mCG
context, less frequently overlapped with all types of TE variants
identified \cite{Schmitz:2013iu}. Furthermore, TE variants were depleted
within genes and DNase I hypersensitivity sites \cite{Sullivan:2014ii},
while they were enriched in gene flanking regions and within other
annotated TEs or pseudogenes (Figure \ref{fig2}B). We found that TE deletions and
common TE variants were enriched within the set of TE variants found in
gene bodies (Figure \ref{fig2}C, D). We did not find any significant enrichment
of TE variants within the \emph{KNOT ENGAGED ELEMENT (KEE)} regions,
previously identified as regions that may act as a ``TE sink''
\cite{Grob:2014kg} (Figure 2 - figure supplement \hyperref[fig2s5]{5}). This may indicate
that these regions do not act as a ``TE sink'' as has been previously
proposed, or that the ``TE sink'' activity is restricted to very recent
insertions, as the insertions we analysed in this population were likely
older than those used in the study by Grob et al.

Among the identified TE variants, several TE superfamilies were over- or
under-represented compared to the number expected by chance given the
overall genomic frequency of different TE types (Figure \ref{fig2}E). In
particular, both TE insertions and deletions in the RC/Helitron
superfamily were less numerous than expected, with an 11.5\% depletion
of RC/Helitron elements in the set of TE variants. In contrast, TEs
belonging to the LTR/Gypsy superfamily were more frequently deleted than
expected, with a 17\% enrichment in the set of TE deletions. This was
unlikely to be due to a differing ability of our detection methods to
identify TE variants of different lengths, as the TE variants identified
had a similar distribution of lengths as all Arabidopsis TEs annotated
in the Col-0 reference genome (Figure 2 - figure supplement \hyperref[fig2s6]{6}). These
enrichments suggest that the RC/Helitron TEs have been relatively
dormant in recent evolutionary history, while the LTR/Gypsy, which are
highly enriched in the pericentromeric regions, are frequently lost from
the Arabidopsis genome. At the family level, we observed similar
patterns of TE variant enrichment or depletion (Figure 2 - figure
supplement \hyperref[fig2s7]{7}; source data 5).

We further examined Arabidopsis (Col-0) DNA sequencing data from a
transgenerational stress experiment to investigate the possible minimum
number of generations required for TE variants to arise
\cite{Jiang:2014ih}. We identified a single potential TE insertion in a
sample following 10 generations of single-seed descent under high
salinity stress conditions, and no TE variants in the control
single-seed descent set. However, without experimental validation it
remains unclear if this represents a true variant. Therefore, we
conclude that TE variants may arise at a rate less than 1 insertion in
30 generations under laboratory conditions. Further experimental work
will be required to precisely determine the rate of transposition in
Arabidopsis.

@@

These results indicate that local DNA methylation patterns are
influenced by the differential TE content between genomes, and that the
DNA methylation-dependent silencing of TEs may lead to formation of DMRs
between wild Arabidopsis accessions. TE insertions appear to be
important in defining local patterns of DNA methylation, while DNA
methylation levels often remain elevated following a TE deletion, and so
are independent from the presence or absence of TEs in these cases.
Importantly, the distance from a TE insertion to the centromere appears
to have a strong impact on whether an alteration of local DNA
methylation patterns will occur. This is likely due to flanking
sequences being highly methylated in the pericentromeric regions, and so
the insertion of a TE cannot further increase levels of DNA methylation.
Overall, a large fraction of the population C-DMRs previously identified
between wild accessions are correlated with the presence of local TE
insertions, but not TE deletions. CG-DMR methylation levels seem to be
mostly independent from the presence/absence of common TEs variants,
while rare TE variants have an impact on DNA methylation levels at both
C-DMRs and CG-DMRs.

\subsection{Genome-wide analysis of TE variant association with methylation
levels highlights distant and local control}

To better quantify the above results, an association scan was conducted
for all common TE variants (\textgreater{}3\% MAF) and all population
C-DMRs for the 124 accessions with both DNA methylation and TE variant
data available. To test the significance of each pairwise correlation,
we collected bootstrap p-value estimates based on 500 permutations of
accession labels. TE-DMR associations were deemed significant if they
had a real association more extreme than any of the 500 permutations (p
\textless{} 1/500). A band of significant associations was observed for
TE insertions and their nearby C-DMRs, signifying a local association
between TE insertion presence/absence and C-DMR methylation (Figure \ref{fig8}A).
This local association was not as strong for TE deletions (Figure \ref{fig8}B),
consistent with our above findings. While TE variants and DNA
methylation showed a local association, it is also possible that TE
variation may influence DNA methylation state more broadly in the
genome, perhaps through production of \emph{trans-}acting smRNAs or
inactivation of genes involved in DNA methylation establishment or
maintenance. To identify any potential enrichment of C-DMRs regulated in
\emph{trans}, we summed the total number of significant associations for
each TE variant across the whole genome (Figure \ref{fig8}A and B, top panels).
At many sites, we found far more significant associations than expected
due to the false positive rate alone. This suggested the existence of
many putative \emph{trans }associations between TE variants and
genome-wide C-DMR methylation levels. We further examined these C-DMRs
that appeared to be associated with a TE insertion in \emph{trans},
checking for TE insertions near these C-DMRs that were present in the
same accessions as the \emph{trans }associated TE, as these could lead
to a false \emph{trans }association. These were extremely rare, with
only 4 such cases for TE insertions, and 38 cases for TE deletions, and
so were unable to explain the high degree of \emph{trans }associations
found.

@@

Here we discovered widespread differential TE content between wild
Arabidopsis accessions, and explore the impact of these variants at the
level of individual accessions. Most TE variants were due to the
\emph{de novo }insertion of TEs, while a smaller subset was likely due
to the deletion of ancestral TE copies, mostly around the
pericentromeric regions. A subset (32\%) of TE variants with a minor
allele frequency above 3\% were able to be tested for linkage with
nearby SNPs. The majority of these TE variants were not in LD with SNPs,
indicating that they represent genetic variants currently overlooked in
genomic studies. We found a marked depletion of TE variants within gene
bodies and DNase I hypersensitivity sites (putative regulatory regions),
indicating that the more deleterious TE insertions may have been removed
from this population through selection. Of those TE variants found in
gene bodies, TE deletions were overrepresented, indicating that the loss
of ancestral TEs inserted within genes may be more frequent, or perhaps
less deleterious, than the \emph{de novo }insertion of TEs into genes.

@@

We ran the TEPID, including the insertion refinement step, on previously
published sequencing data for 216 different Arabidopsis populations
(NCBI SRA SRA012474; Schmitz et al. 2013), mapping to the TAIR10
reference genome and using the TAIR9 TE annotation. The `--mask' option
was also used to mask the mitochondrial and plastid genomes. We also ran
TEPID using previously published transgenerational data for salt stress
and control conditions (NCBI SRA SRP045804; \cite{Jiang:2014ih}), using
the `--mask' option to mask mitochondrial and plastid genomes, and the
`--strict' option for highly related samples.

@@

Minor allele frequency distribution for TE variants.

>>>>

\usepackage[scaled]{helvet}
\renewcommand\familydefault{\sfdefault}
\usepackage[T1]{fontenc}

@@

\renewcommand\Authfont{\fontsize{13}{14.4}\selectfont}
\renewcommand\Affilfont{\fontsize{10}{10.8}\selectfont}

@@

Variation in the presence or absence of transposable elements (TEs) is
a major source of genetic variation between individuals. Here, we
identified 23,095 TE presence/absence variants between 216 Arabidopsis
accessions. Most TE variants were rare, and we find a burden of rare
variants associated with local extremes of both gene expression and
DNA methylation levels within the population. Of the common alleles
identified, two thirds were not in linkage disequilibrium with nearby
SNPs, implicating these TE variants as a source of novel genetic
diversity. Nearly 200 common TE variants were associated with
significantly altered expression of nearby genes, and a major fraction
of inter-accession DNA methylation differences were associated with
nearby TE insertion variants. Overall, we find that TE variants likely
play an important role in facilitating epigenomic and transcriptional
variation between wild Arabidopsis accessions, indicating a strong
genetic basis for epigenetic variation.

@@

Transposable elements (TEs) are mobile genetic elements present in
nearly all studied organisms, and comprise a large fraction of most
eukaryotic genomes. The two classes of TEs are retrotransposons (class
I elements), which transpose via an RNA intermediate requiring a
reverse transcription reaction, and DNA transposons (class II
elements), which transpose via either a cut-paste or, in the case of
Helitrons, a rolling circle mechanism with no RNA intermediate
\cite{Wicker:2007en}. TE activity poses mutagenic potential as a TE
insertion may disrupt functional regions of the genome. Consequently,
safeguard mechanisms have evolved to suppress this activity, including
the methylation of cytosine nucleotides (DNA methylation) to produce
5-methylcytosine (mC), a modification that can induce transcriptional
silencing of the methylated locus. In \emph{Arabidopsis thaliana}
(Arabidopsis), DNA methylation occurs in all three DNA sequence
contexts: mCG, mCHG, and mCHH, where H is any base but
G. Establishment of DNA methylation marks can be carried out by two
distinct pathways -- the RNA-directed DNA methylation pathway guided
by 24 nucleotide (nt) small RNAs (smRNAs), and the DDM1/CMT2 pathway
\cite{Zemach:2013dj, Matzke:2014ek}. A major function of DNA
methylation in Arabidopsis is in the transcriptional silencing of TEs.
Loss of DNA methylation due to mutations in genes essential for its
establishment or maintenance leads to expression of previously silent
TEs, and in some cases transposition \cite{Mirouze:2009km,Miura:2001eg, Saze:2003du, Lippman:2004cm, Jeddeloh:1999gu,Zemach:2013dj}.

TEs are thought to play an important role in evolution, not only because
of the disruptive potential of their transposition. The release of
transcriptional and post-transcriptional silencing of TEs can lead to
bursts of TE activity, rapidly generating new genetic diversity
\cite{Vitte:2014he}. TEs may carry regulatory information such as
promoters and transcription factor binding sites, and their mobilization
may lead to the creation or expansion of gene regulatory networks
\cite{Henaff:2014hg, Bolger:2014bn, Ito:2011dga, Makarevitch:2015ho}.
Furthermore, the transposase enzymes required and encoded by TEs have
frequently been domesticated and repurposed as endogenous proteins, such
as the \emph{DAYSLEEPER} gene in Arabidopsis, derived from a hAT
transposase enzyme \cite{Bundock:2005gp}. Clearly, the activity of TEs
can have widespread and unpredictable effects on the host genome.
However, the identification of TE presence/absence variants in genomes
has remained difficult to date. It is challenging to identify the
structural changes in the genome caused by TE mobilization using current
short-read sequencing technologies as these reads are typically mapped
to a reference genome, which has the effect of masking structural
changes that may be present. However, in terms of the number of base
pairs affected, a large fraction of genetic differences between
Arabidopsis accessions appears to be due to variation in TE content
\cite{Cao:2011cf, Quadrana:2016bi}. Therefore identification of TE
variants is essential in order to develop a more comprehensive
understanding of the genetic variation that exists between genomes, and
of the consequences of TE movement on genome and cellular function.

The tools developed previously for identification of novel TE insertion
events have several limitations. They either require a library of active
TE sequences, cannot identify TE absence variants, are not designed
with population studies in mind, or suffer from a high degree of
false-negatives
\cite{Thung:2014ir, Robb:2013gw, Henaff:2015dl, Quadrana:2016bi}. In
order to accurately map the locations of TE presence/absence variants
with respect to a reference genome, we have developed a novel algorithm,
TEPID (Transposable Element Polymorphism IDentification), which is
designed for population studies. We tested our algorithm using both
simulated and real Arabidopsis sequencing data, finding that TEPID is
able to accurately identify TE presence/absence variants with respect to
the Col-0 reference genome. We applied our TE analysis method to
existing genome resequencing data for 216 different wild Arabidopsis
accessions, and identified widespread TE variation amongst these
accessions \cite{Schmitz:2013iu}.

@@

In order to further assess the sensitivity and specificity of TE variant
discovery using TEPID, we identified TE variants in the Landsberg
\emph{erecta} (L\emph{er}) accession, and compared these with the
L\emph{er }genome assembly created using long PacBio sequencing reads
\cite{Chin:2013iw}. Previously published 100 bp paired-end L\emph{er}
genome resequencing reads \cite{Schneeberger:2011ft} were first analyzed
using TEPID, enabling identification of 446 TE insertions (Figure 1 -
source data 1) and 758 TE absence variants (Figure 1 - source data 2)
with respect to the Col-0 reference TE annotation. Reads providing
evidence for these variants were then mapped to the L\emph{er} reference
genome, generated by \emph{de novo} assembly using Pacific
Biosciences P5-C3 chemistry with a 20 kb insert library
\cite{Chin:2013iw}, using the same alignment parameters as was used to
map reads to the Col-0 reference genome. This resulted in 98.7\% of
reads being aligned concordantly to the L\emph{er} reference, whereas
100\% aligned discordantly or as split reads to the Col-0 reference
genome (Table \ref{table1}). To find whether reads mapped to homologous regions in
both the Col-0 and L\emph{er} reference genomes, we conducted a blast
search \cite{Camacho:2009fc} using the DNA sequence between read pair
mapping locations in the L\emph{er }genome against the Col-0 genome, and
found the top blast result for 80\% of reads providing evidence for TE
insertions, and 89\% of reads providing evidence for TE absence variants
in L\emph{er}, to be located within 200 bp of the TE variant
reported by TEPID. We conclude that reads providing evidence for TE
variants map discordantly or as split reads when mapped to the Col-0
reference genome, but map concordantly to homologous regions of the
L\emph{er} \emph{de novo} assembled reference genome, indicating that
structural variation is present at the sites identified by TEPID, and
that this is resolved in the \emph{de novo} assembled genome.

To estimate the rate of false negative TE absence calls made using
TEPID, we compared our L\emph{er} TE absence calls to the set of TE
absences in L\emph{er} genome identified previously by aligning
full-length Col-0 TEs to the L\emph{er} reference using BLAT
\cite{Quadrana:2016bi}. We found that 89.6\% (173/193) of these TE
absences were also identified using TEPID, indicating a false negative
rate of around 10\% for TE absence calls. To determine the rate of false
negative TE insertion calls, we ran TEPID using 90 bp paired-end Col-0
reads (Col-0 control samples from \cite{Jiang:2014ih}), aligning reads
to the L\emph{er} PacBio assembly. As TEPID requires a high-quality TE
annotation to discover TE variants, which is not available for the
L\emph{er} assembly, we simply looked for discordant and split read
evidence at the known Col-0-specific TE insertion sites
\cite{Quadrana:2016bi}, and found evidence reaching the TEPID threshold
for a TE insertion call to be made at 89.6\% (173/193) of these sites,
indicating a false negative rate of 10\%. However, we note that this
estimate does not take into account the TEPID refinement step used on
large populations, and so the false negative rate for samples analyzed
in the population from Schmitz et al. (2013) is likely to be slightly
lower than this estimate, as each accession gained on average 4\% more
insertion calls following this refinement step (Figure 2 - figure
supplement \hyperref[fig2s1]{1}).

@@

We used TEPID to analyze previously published 100 bp paired-end genome
resequencing data for 216 different Arabidopsis accessions
\cite{Schmitz:2013iu}, and identified 15,007 TE insertions (Figure 2 -
source data 1) and 8,088 TE absence variants (Figure 2 - source data
2) relative to the Col-0 reference accession, totalling 23,095 unique
TE variants. In most accessions we identified 300-500 TE insertions
(mean = 378) and 1,000-1,500 TE absence variants (mean = 1,279), the
majority of which were shared by two or more accessions (Figure 2 -
figure supplement \hyperref[fig2s2]{2}). PCR validations were
performed for a random subset of 10 insertions and 10 absence variants
in 14 accessions (totalling 280 validations), and confirmed the high
accuracy of TE variant discovery using the TEPID package, with results
similar to that observed using simulated data and the L\emph{er}
genome analysis (Figure 2 - figure supplement
\hyperref[fig2s3]{3}). The number of TE insertions identified was
positively correlated with sequencing depth of coverage, while the
number of TE absence variants identified had no correlation with
sequencing coverage (Figure 2 - figure supplement
\hyperref[fig2s4]{4}A, B), indicating that the sensitivity of TE
absence calls is not limited by sequencing depth, while TE insertion
calls benefitted from high sequencing depth.  However, accessions with
low coverage gained more TE insertion calls during the TEPID
refinement step (Figure 2 - figure supplement \hyperref[fig2s4]{4}C),
indicating that these false negatives were effectively reduced by
leveraging TE variant information for the whole population.

As TE insertion and TE absence calls represent an arbitrary comparison
to the Col-0 reference genome, we sought to remove these arbitrary
comparisons and classify each variant as a new TE insertion or true
deletion of an ancestral TE in the population. To do this, we examined
the minor allele frequency (MAF) of each variant in the population,
under the expectation that the minor allele is the derived allele. We
re-classified common TE absences relative to Col-0 as rare TE
insertions in Col-0, and common TE insertions relative to Col-0 as
true TE deletions in Col-0. Cases where the TE variant was at an high
MAF (\textgreater{}20\%) were assigned NA calls, as it could not be
determined if these were cases where the variant was most likely to be
due to a true TE deletion or due to a new TE insertion. While these
classifications are not definitive, as there will be rare cases where
a true TE deletion has spread through the population and becomes the
common allele, we will correctly classify most TE variants. Overall,
we found 72.3\% of the TE absence variants identified with respect to
the Col-0 reference genome were likely due to a true TE deletion in
these accessions, while 4.8\% were due to new insertions in Col-0 not
shared by others in the population (Table \ref{table2}). Overall, we
identified 15,077 new TE insertions, 5,856 true TE deletions, and
2,162 TE variants at a high MAF that were unable to be classified as
an insertion or deletion (Figure 2 - source data 3).

TE insertions and deletions were distributed throughout chromosome 1 in
a pattern that was similar to the distribution of all Col-0 TEs (Figure
\ref{fig2}A). TE deletions and common TE variants were found in similar
chromosomal regions, as deletion variants represent the rare loss of
common variants. TE deletions and common variants were more highly
enriched in the pericentromeric regions than rare variants or TE
insertions. The distribution of rare TE variants and TE insertions was
somewhat similar to that observed for regions of the genome previously
identified as being differentially methylated in all DNA methylation
contexts (mCG, mCHG, mCHH) between the wild accessions (population
C-DMRs), while population CG-DMRs, differentially methylated in the mCG
context, less frequently overlapped with all types of TE variants
identified \cite{Schmitz:2013iu}. Furthermore, TE variants were depleted
within genes and DNase I hypersensitivity sites \cite{Sullivan:2014ii},
while they were enriched in gene flanking regions and within other
annotated TEs or pseudogenes (Figure \ref{fig2}B). We found that TE deletions and
common TE variants were enriched within the set of TE variants found in
gene bodies (Figure \ref{fig2}C, D). We did not find any significant enrichment
of TE variants within the \emph{KNOT ENGAGED ELEMENT (KEE)} regions,
previously identified as regions that may act as a ``TE sink''
\cite{Grob:2014kg} (Figure 2 - figure supplement \hyperref[fig2s5]{5}). This may indicate
that these regions do not act as a ``TE sink'' as has been previously
proposed, or that the ``TE sink'' activity is restricted to very recent
insertions, as the insertions we analysed in this population were likely
older than those used in the \emph{KEE} study \cite{Grob:2014kg}.

Among the identified TE variants, several TE superfamilies were over-
or under-represented compared to the number expected by chance given
the overall genomic frequency of different TE types (Figure
\ref{fig2}E). In particular, both TE insertions and deletions in the
RC/Helitron superfamily were less numerous than expected, with an
11.5\% depletion of RC/Helitron elements in the set of TE variants. In
contrast, TEs belonging to the LTR/Gypsy superfamily were more
frequently deleted than expected, with a 17\% enrichment in the set of
TE deletions. This was unlikely to be due to a differing ability of
our detection methods to identify TE variants of different lengths, as
the TE variants identified had a similar distribution of lengths as
all Arabidopsis TEs annotated in the Col-0 reference genome (Figure 2
- figure supplement \hyperref[fig2s6]{6}). These enrichments suggest
that the RC/Helitron TEs have been relatively dormant in recent
evolutionary history, while the LTR/Gypsy, which are highly enriched
in the pericentromeric regions, are frequently lost from the
Arabidopsis genome. At the family level, we observed similar patterns
of TE variant enrichment or depletion (Figure 2 - figure supplement
\hyperref[fig2s7]{7}; source data 4).

We further examined Arabidopsis (Col-0) DNA sequencing data from a
transgenerational stress experiment to investigate the possible
minimum number of generations required for TE variants to arise
\cite{Jiang:2014ih}. In one of the three replicates, we identified a
single potential TE insertion in a sample following 10 generations of
single-seed descent under high salinity stress conditions, and no TE
variants in the control single-seed descent set. However, without
experimental validation it remains unclear if this represents a true
variant. Therefore, we conclude that TE variants may arise at a rate
less than 1 insertion in 30 generations under laboratory
conditions. Further experimental work will be required to precisely
determine the rate of transposition in Arabidopsis.

@@

These results indicate that local DNA methylation patterns are
influenced by the differential TE content between genomes, and that the
DNA methylation-dependent silencing of TEs may lead to formation of DMRs
between wild Arabidopsis accessions. TE insertions appear to be
important in defining local patterns of DNA methylation, while DNA
methylation levels often remain elevated following a TE deletion, and so
are independent from the presence or absence of TEs in these cases.
Importantly, the distance from a TE insertion to the centromere appears
to have a strong impact on whether an alteration of local DNA
methylation patterns will occur. This is likely due to flanking
sequences being highly methylated in the pericentromeric regions, and so
the insertion of a TE cannot further increase levels of DNA methylation.
Overall, a large fraction of the population C-DMRs previously identified
between wild accessions are correlated with the presence of local TE
insertions, but not TE deletions. CG-DMR methylation levels seem to be
mostly independent from the presence/absence of common TE variants,
while rare TE variants have an impact on DNA methylation levels at both
C-DMRs and CG-DMRs.

\subsection{Genome-wide association scan highlights
  distant and local control of DNA methylation}

To better quantify the above results, an association scan was conducted
for all common TE variants (\textgreater{}3\% MAF) and all population
C-DMRs for the 124 accessions with both DNA methylation and TE variant
data available. To test the significance of each pairwise correlation,
we collected bootstrap p-value estimates based on 500 permutations of
accession labels. TE-DMR associations were deemed significant if they
had an association more extreme than any of the 500 permutations (p
\textless{} 1/500). A band of significant associations was observed for
TE insertions and their nearby C-DMRs, signifying a local association
between TE insertion presence/absence and C-DMR methylation (Figure \ref{fig8}A).
This local association was not as strong for TE deletions (Figure \ref{fig8}B),
consistent with our above findings. While TE variants and DNA
methylation showed a local association, it is also possible that TE
variation may influence DNA methylation state more broadly in the
genome, perhaps through production of \emph{trans-}acting smRNAs or
inactivation of genes involved in DNA methylation establishment or
maintenance. To identify any potential enrichment of C-DMRs regulated in
\emph{trans}, we summed the total number of significant associations for
each TE variant across the whole genome (Figure \ref{fig8}A and B, top panels).
At many sites, we found far more significant associations than expected
due to the false positive rate alone. This suggested the existence of
many putative \emph{trans }associations between TE variants and
genome-wide C-DMR methylation levels. We further examined these C-DMRs
that appeared to be associated with a TE insertion in \emph{trans},
checking for TE insertions near these C-DMRs that were present in the
same accessions as the \emph{trans }associated TE, as these could lead
to a false \emph{trans }association. These were extremely rare, with
only 4 such cases for TE insertions, and 38 cases for TE deletions, and
so were unable to explain the high degree of \emph{trans }associations
found.

@@

Here we discovered widespread differential TE content between wild
Arabidopsis accessions, and explore the impact of these variants at the
level of individual accessions. Most TE variants were due to the
\emph{de novo }insertion of TEs, while a smaller subset was likely due
to the deletion of ancestral TE copies, mostly around the
pericentromeric regions. A subset (32\%) of TE variants with a minor
allele frequency above 3\% were able to be tested for linkage with
nearby SNPs. The majority of these TE variants were not in LD with SNPs,
indicating that they represent genetic variants currently overlooked in
genomic studies. We found a marked depletion of TE variants within gene
bodies and DNase I hypersensitivity sites (putative regulatory regions),
consistent with the more deleterious TE insertions have being removed
from this population through selection. Of those TE variants found in
gene bodies, TE deletions were overrepresented, indicating that the loss
of ancestral TEs inserted within genes may be more frequent, or perhaps
less deleterious, than the \emph{de novo }insertion of TEs into genes.

@@

We ran TEPID, including the insertion refinement step, on previously
published sequencing data for 216 different Arabidopsis populations
(NCBI SRA SRA012474; \cite{Schmitz:2013iu}), mapping to the TAIR10
reference genome and using the TAIR9 TE annotation. The `--mask' option
was set to mask the mitochondrial and plastid genomes. We also ran
TEPID using previously published transgenerational data for salt stress
and control conditions (NCBI SRA SRP045804; \cite{Jiang:2014ih}), again using
the `--mask' option to mask mitochondrial and plastid genomes, and the
`--strict' option for highly related samples.

@@

Number of accessions sharing TE variants identified by TEPID.
